Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2016-07-10 RMS/submission.tex
modified a bit

In this paper, we only consider Real-Time Maude's \emph{Linear
  Temporal Logic (LTL) model checker}, which analyzes whether
\emph{each} behavior satisfies a temporal logic formula. \emph{State
  propositions} are defined as terms of sort \verb|Prop|. Their
semantics is defined by conditional equations of the form $\texttt{ceq
} statePattern \texttt{ |= } prop \texttt{ = } b \texttt{ if } cond$,
with $b$ a term of sort \verb|Bool|, stating that $prop$ evaluates to
$b$ in states which are instances of $statePattern$ provided the
condition $cond$ holds. These equations together define $prop$ to hold
in all states $s$ that make $s \texttt{ |= } prop$ evaluate to
\verb|true|. A temporal logic \emph{formula} is constructed by state
propositions and temporal logic operators such as \verb|True|,
\verb|False|, \verb|~|(negation), \verb|/\|, \verb|\/|,
\verb|->|(implication), \verb|[]|(``always'') and
\verb|U|(``until''). Real-Time Maude supports both \emph{timed} and
\emph{untimed LTL model checking}. The untimed model checking command
\begin{alltt}
  (mc \(s\) |=u \(\Phi\) .)
\end{alltt}
checks whether the temporal logic formula $\Phi$ holds in all
behaviors starting from the initial state $s$, \emph{with no time
  limit}.

@@

To verify the schedulability of a given set of periodic tasks, we
define an atomic proposition \verb|taskTimeout| to hold if there
exists an \verb|error| in $taskList$ of the current state, that is,
some task misses its deadline:
\begin{verbatim}
  op taskTimeout : -> Prop [ctor] .
  eq {L T STS HW ISRC} |= taskTimeout 
     = containError(L) .
\end{verbatim}
where \verb|containError| returns \verb|true| if there is an
\verb|error| existing in \verb|L|. Then schedulability can be
formalized as the temporal logic formula: \verb|[](~taskTimeout)|. As
the property is not \emph{clock-related}, given an initial state
\verb|init|, the following untimed model checking command returns
\verb|true| if the schedulability property holds with no time limit;
otherwise a trace showing a counterexample is provided:
\begin{verbatim}
  (mc init |=u [](~taskTimeout) .)
\end{verbatim}

Another important objective is to verify the correctness of the
implementation.  The atomic proposition \verb|correct| is hence
defined to hold if the running periodic task is the one requested to
be executed with the highest priority:
\begin{verbatim}
  op correct : -> Prop [ctor] .
  ceq {L T STS HW ISRC} |= correct
      = if ID :: MaybeNat then shouldRun(ID, L)
        else true fi
      if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is not
\verb|DORMANT|. Note that during verification, we do not care the
behaviors after some task misses its deadline. Therefore, the
correctness property is formalized by the temporal logic formula:
\verb|([]correct) \/ (correct U taskTimeout)|, and can be verified by
the following untimed model checking command provided an initial state
\verb|init|:
\begin{verbatim}
  (mc init |=u ([]correct) 
               \/ (correct U taskTimeout) .)
\end{verbatim}

>>>>

In this paper, we only consider Real-Time Maude's \emph{Linear
  Temporal Logic (LTL) model checker}, which analyzes whether
\emph{each} behavior satisfies a temporal logic formula. \emph{State
  propositions} are defined as terms of sort \verb|Prop|. Their
semantics is defined by conditional equations of the form $\texttt{ceq
} statePattern \texttt{ |= } prop \texttt{ = } b \texttt{ if } cond$,
with $b$ a term of sort \verb|Bool|, stating that $prop$ evaluates to
$b$ in states which are instances of $statePattern$ provided the
condition $cond$ holds. These equations together define $prop$ to hold
in all states $s$ that make $s \texttt{ |= } prop$ evaluate to
\verb|true|. A temporal logic \emph{formula} is constructed by state
propositions and temporal logic operators such as \verb|~|(negation),
\verb|\/|(or), \verb|[]|(``always'') and
\verb|U|(``until''). Real-Time Maude supports both \emph{timed} and
\emph{untimed LTL model checking}. The untimed model checking command
\begin{alltt}
  (mc \(s\) |=u \(\Phi\) .)
\end{alltt}
checks whether the temporal logic formula $\Phi$ holds in all
behaviors starting from the initial state $s$, \emph{with no time
  limit}.

@@

To verify the schedulability of a given set of periodic tasks, we
define an atomic proposition \verb|taskTimeout| to hold if there
exists an \verb|error| in $taskList$ of the current state, that is,
some task misses its deadline:
\begin{verbatim}
  op taskTimeout : -> Prop [ctor] .
  eq {L T STS HW ISRC} |= taskTimeout 
     = containError(L) .
\end{verbatim}
where \verb|containError| returns \verb|true| if there is an
\verb|error| existing in \verb|L|. Then schedulability can be
formalized as the temporal logic formula: \verb|[](~taskTimeout)|,
\hl{expressing that the proposition} \verb|taskTimeout| \hl{is always false}. As
the property is not \emph{clock-related}, given an initial state
\verb|init|, the following untimed model checking command returns
\verb|true| if the schedulability property holds with no time limit;
otherwise a trace showing a counterexample is provided:
\begin{verbatim}
  (mc init |=u [](~taskTimeout) .)
\end{verbatim}

Another important objective is to verify the correctness of the
implementation.  The atomic proposition \verb|correct| is hence
defined to hold if the running periodic task is the one requested to
be executed with the highest priority:
\begin{verbatim}
  op correct : -> Prop [ctor] .
  ceq {L T STS HW ISRC} |= correct
      = if ID :: MaybeNat then shouldRun(ID, L)
        else true fi
      if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is not
\verb|DORMANT|. Note that during verification, we do not care the
behaviors after some task misses its deadline. Therefore, the
correctness property is formalized by the temporal logic formula:
\verb|([]correct)\/(correct U taskTimeout)|, \hl{stating that}
\verb|correct| \hl{is always true, or is true until} \verb|taskTimeout|
\hl{holds}. It can be verified by the following untimed model checking
command provided an initial state \verb|init|:
\begin{verbatim}
  (mc init |=u ([]correct) 
               \/ (correct U taskTimeout) .)
\end{verbatim}
